---
title: 'Final Project'
subtitle: "PSTAT122: Design and Analysis of Experiments" 

author:
  - name: "Fall 2025"
   # affiliations:
    #  - name: "Fall 2025"
affiliation-title: "Quarter"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.2in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)

```

:::: callout
::: {style="text-align: center"}
[**STUDENT NAME**]{style="color: blue;"}
:::

-   STUDENT 1 (NetID 1)
-   STUDENT 2 (NetID 2)
-   STUDENT 3 (NetID 3)
-   STUDENT 4 (NetID 4)
::::

::: callout-caution
# Due Date

**Due Date:** Monday, December 8, 2025, 11:59 PM
:::

# Introduction

-   Clear statement of the objective or research question.
-   Brief context or motivation.

# Experimental Design

-   Description of factors and treatment structure.
-   Clearly state what you are measuring and the units. Examples: Number of words recalled (count), reaction time (seconds), taste rating (1–5 scale).
-   Identify which factors are fixed vs. random.
-   Description of design type (CRD, RCBD, factorial, etc.).
-   Explain how randomization, replication, and (if used) blocking were implemented.
-   Sample size: Provide number of observations per condition. Guideline: 5–10 per treatment for CRD, 3–5 blocks for RCBD, total feasible within 1 hour.

(You are encouraged to explore more resources for determining the sample size )

# Data Collection

-   **Procedure**: Describe how and when the experiment was conducted (e.g., location, date, steps taken).
-   **Challenges/Adjustments**: Mention any difficulties or changes made during data collection (e.g., technical issues, time adjustments).
-   **Data Presentation**: Display the collected data in tables or graphs, summarizing key measures like mean and standard deviation..\

```{r, echo = FALSE}
set.seed(123)

participants <- c("Chris", "Christian", "Isaac", "Tommy")
conditions <- data.frame(
  Music = c("No", "Yes", "No", "Yes"),
  Gum   = c("No", "No", "Yes", "Yes")
)


randomize_conditions <- function(name) {
  cond <- conditions[sample(1:4), ]
  cond$Participant <- name
  cond
}

design_list <- lapply(sample(participants), randomize_conditions)
design_df <- do.call(rbind, design_list)

design_df$Run1 <- c(62, 66, 60, 65, 57, 91, 94, 82, 78, 82, 82, 82, 60, 66, 72, 62)
design_df$Run2 <- c(51, 61, 56, 62, 51, 65, 86, 84, 79, 76, 91, 82, 66, 65, 49, 65)
design_df$Run3 <- c(54, 53, 46, 58, 54, 94, 81, 82, 96, 69, 87, 74, 58, 58, 80, 67)
design_df$Run4 <- c(55, 67, 54, 60, 84, 76, 81, 83, 90, 86, 88, 84, 60, 65, 70, 62)
design_df$Run5 <- c(63, 62, 58, 61, 79, 76, 79, 81, 89, 91, 97, 88, 64, 70, 56, 67)
design_df$Run6 <- c(61, 62, 55, 60, 90, 88, 74, 74, 82, 85, 79, 87, 56, 67, 59, 67)
design_df$Run7 <- c(60, 63, 64, 60, 68, 69, 71, 80, 97, 90, 87, 70, 62, 62, 58, 63)
design_df$Run8 <- c(64, 62, 73, 66, 81, 78, 78, 77, 93, 58, 84, 81, 59, 44, 56, 56)
design_df$Run9 <- c(54, 66, 59, 60, 82, 86, 86, 83, 90, 65, 84, 74, 66, 67, 64, 67)
design_df$Run10 <- c(60, 66, 66, 65, 87, 76, 70, 84, 90, 83, 80, 70, 72, 66, 60, 71)
design_df$Avg <- c(59.4, 62.1, 68, 59.6, 91.3, 79.9, 78, 76.8, 98.1, 78.6, 88.9, 81, 62.5, 62, 90.4, 63.7)

design_df <- design_df[, c("Participant", "Music", "Gum",
                           "Run1", "Run2", "Run3", "Run4", "Run5", "Run6", "Run7", "Run8","Run9","Run10","Avg")]


knitr::kable(design_df)

```

**Christian Solares: Data frame:**

```{r}
# 1) DATAFRAME:
monkey_data_frame <- data.frame(
    participants = rep(c("Isaac", "Tommy", "Chris", "Christian"), each = 4) ,
    
    music = c("No", "Yes", "Yes", "No", 
              "No", "No", "Yes", "Yes",
              "Yes", "No", "Yes", "No",
              "Yes", "No", "No", "Yes") ,
  
    gum = c("Yes", "No", "Yes", "No",
              "Yes", "No", "No", "Yes",
              "No", "No", "Yes", "Yes",
              "Yes", "No", "Yes", "No") ,
    
    run1 = c(62, 66, 60, 65, 57, 91, 94, 82, 78, 82, 82, 82, 60, 66, 72, 62),
    run2 = c(51, 61, 56, 62, 51, 65, 86, 84, 79, 76, 91, 82, 66, 65, 49, 65),
    run3 = c(54, 53, 46, 58, 54, 94, 81, 82, 96, 69, 87, 74, 58, 58, 80, 67),
    run4 = c(55, 67, 54, 60, 84, 76, 81, 83, 90, 86, 88, 84, 60, 65, 70, 62),
    run5 = c(63, 62, 58, 61, 79, 76, 79, 81, 89, 91, 97, 88, 64, 70, 56, 67),
    run6 = c(61, 62, 55, 60, 90, 88, 74, 74, 82, 85, 79, 87, 56, 67, 59, 67),
    run7 = c(60, 63, 64, 60, 68, 69, 71, 80, 97, 90, 87, 70, 62, 62, 58, 63),
    run8 = c(64, 62, 73, 66, 81, 78, 78, 77, 93, 58, 84, 81, 59, 44, 56, 56),
    run9 = c(54, 66, 59, 60, 82, 86, 86, 83, 90, 65, 84, 74, 66, 67, 64, 67),
    run10 = c(60, 66, 66, 65, 87, 76, 70, 84, 90, 83, 80, 70, 72, 66, 60, 71),
    average = c(59.4, 62.1, 68, 59.6, 91.3, 79.9, 78, 76.8, 98.1, 78.6, 88.9, 81, 62.5, 62, 90.4, 63.7)
)

monkey_data_frame
# Factorial ANOVA with blocking
# gum + music + participants (looking at the effect of gum, music, and participants seperately) 
# gum*msuic (interaction between gum and music)
model <- lm(average ~ gum + music + gum*music + participants, data = monkey_data_frame)
anova_result <- anova(model)
anova_result


summary_result <- summary(model)
print(summary_result)

# 4) Visuals Of Our Data:


```

\

# Analysis

-   **Exploratory Data**: Start with basic statistics (mean, SD) and visualizations (e.g., boxplots) to understand the data.

-   **Hypothesis Testing**: Test your hypothesis with an appropriate statistical test (e.g., ANOVA).

-   **Tables, Figures, & Code**: Include key results (ANOVA table, post-test) and relevant `R` code excerpts where needed.

-   Use `R` to analyze the data.

# Conclusions

-   Summarize key findings.
-   Comment on limitations and possible improvements.

# References

(If needed.)

# Appendices

-   `R` code.
